---
title: "GWASr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GWASr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
author:
  - Franjo Ivankovic^[University of Florida, njofrica@ufl.edu]
  - Mingjing Chen^[University of Florida, mingjingchen@ufl.edu]
  - Xiaoxiao Wen^[University of Florida, xiaoxiao.wen@ufl.edu]
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

GWASr is a simple R pacakge which is capable of building QQ and Manhattan plots as well as calculating genomic lambda constant.

## Install Packages
To install GWASr, a simple one-line code below with fetch this package from GitHub and install it in R. Additionally, the *build_vignettes = TRUE* option will automatically build an HTML of this vignette.
```{r install}
devtools::install_github("FranjoIM/PackageName", ref = "v0.1.0", build_vignettes = TRUE)
```

## Load the Package
Loading library is very easy, by simply running the following one-line code.
```{r setup}
library(GWASr)
```

## Accessing Vignettes
To access this vignette, simply run the following code in the R console.
```{r vignette}
vignette("GWASr", package = "GWASr")
```

## Overview of GWASr Functions
GWASr includes 3 functions, namely 2 plottng functions written in R (*plot_manhattan* and *plot_qq*), and 1 function written in Rcpp (*GLambda*). While the main focus of the package are the plotting functions, we will overview the *GLambda* first, as it's fairly straight forward. 

### Genomic Lambda Calculation (Rcpp)
Genomic lambda is a numerical constant, which is defined as the ratio of the median $\chi^2$ test score to an $\chi^2$ at $p=0.5$, or $0.4549364$. Genomic Lambda, also called the genomic inflation factor $\lambda_{gc}$, measures systematic biases that may be present in the GWAS association results. The expected $\lambda_{gc} \leq 1$. If the $\lambda_{gc} > 1$, then this may be evidence for some systematic bias that needs to be corrected in the analysis. 

The easiest way to calculate $\lambda_{gc}$ is to pass the p values from the summary statistics directly into the function. And example using the included data is:
```{r Genomic lambda}
GLambda(ADHDMetaP$P)
```

Any numeric variable can be passed into the function. Additionally, by the very mathematical definition provided above, the $\lambda_{gc}$ when a single $p = 0.5$ is passed in should evaluate to $1$.
```{r Genomic lambda tes}
GLambda(0.5)
#> 1
```

For a detailed overview of arguments, run:
```{r Geenomic lambda arguments}
?GLambda
```

### Manhattan Plot
Manhattan plot is the most frequent method of visually representing the strength of genome-wide associations, by plotting each SNP's association $p$ on a y-axis, specifically as a $-log_{10}$ transform thereof, while plotting genomic positions on the x-axis. Genomic positions are mapped by typical *chr: base* address. Additionally, this function plots two reference lines on the plot at $p_{stringent} = 3*10^-8$ and $p_{lax} = 10^-5$, which represent genome-wide significant and genome-wide suggestive thresholds, respectively.

This function requires a data frame with 3 columns: chromosome ID (character class), base-pair position (numeric class), and $p$ values (numeric types). 
```{r Manhattan plot}
plot_manhattan(ADHDMeta)
```

For a detailed overview of arguments, run:
```{r Manhattan plot arguments}
?plot_manhattan
```

### QQ Plot
In a genome-wide association study, most of the SNPs tested should not to be associated with the phenotype. Thus, ideally, qq-plot of GWAS $p$ should follow a straight diagonal line with expected $p$, only taperig off at the very end. QQ plots show the quantile distribution of the $p$ values from our samples. QQ plots can inform whether there is a presence of systematic bias, much like the genomic lambda, and indicate if a closer look at data and model are necessary.

This function requres a data frame with 1 column: $p$ values (numeric types).
```{r QQ plot}
plot_qq(DF)
```

For a detailed overview of arguments, run:
```{r QQ plot arguments}
?plot_qq
```

## Included Datasets
This package includes two datasets which were obtained from [Psychiatric Genomics Consortium](https://www.med.unc.edu/pgc/download-results/) results page, and were originally published by [Demontis et al. (2019)](https://doi.org/10.1038/s41588-018-0269-7) in *Nature Genetics*.

## ADHD Meta
ADHD Meta is a processed data frame of 239,961 rows and 5 variables. In order to reduce memory requirements for packaging and distribution, as well as running trial functions, all observations with $p$ values over $0.2$ were excluded in the set.

The variables include:

1. CHR<chr>: chromosome ID
2. SNP<chr>: SNP ID
3. BP<num>: base-pair genomic location
4. OR<num>: Odds-ratio from GWAS
5. P<num>: P values from GWAS

As lazy-loading is on for this package, the data should load with the package and callable ass:
```{r ADHDMeta}
ADHDMeta
names(ADHDMeta)
#> [1] "CHR" "SNP" "BP"  "OR"  "P"  
```

## ADHD Meta P
ADHD Meta P is a minimally processed collection of SNP IDs and related $p$ values. It includes 8,094,094 rows and 2 variables. Most of the columns were removed in thee interest of reducing memory requirements, but all the $p$ values were kept as reported in the study.

The variables include:

1. SNP<chr>: SNP ID
2. P<num>: P values from GWAS

As lazy-loading is on for this package, the data should load with the package and callable ass:
```{r ADHDMetaP}
ADHDMetaP
names(ADHDMetaP)
#> [1] "SNP" "P"  
```

## References
* Demontis, D., Walters, R.K., Martin, J. et al. Discovery of the first genome-wide significant risk loci for attention deficit/hyperactivity disorder. *Nat Genet* 51, 63–75 (2019). [10.1038/s41588-018-0269-7](https://doi.org/10.1038/s41588-018-0269-7)
